---
title: "R_Markdown_Valerie"
author: "Valerie Yu"
date: "`r Sys.Date()`"
output: pdf_document
---

# Valerie's R Assignment
Setting up the R Markdown and loading in R library.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())
library(tidyverse)
library(ggplot2)
library(tidyr)
library(tibble)
library(dplyr)
```


## Part I
### _Data Inspection_
`fang_et_al_genotypes.txt`

* There are 2782 rows and 986 columns.
* The first 3 columns are `Sample_ID`, `JG_OTU`, and `Group`. Then the rest are SNP data (983).

`snp_position.txt`

* There are 983 rows and 15 columns.
* The columns are named `SNP_ID`, `Chromosome`, `Position`, etc.

```{r}
# Loading in data from "data" folder and looking at column names. 
genotypes <- read_tsv("data/fang_et_al_genotypes.txt")
snp_position <- read_tsv("data/snp_position.txt")


# Inspecting the data
dim(genotypes) # 2782 rows  986 columns
colnames(genotypes)[1:15] # variable names 
# head(genotypes) # 1st Sample_ID, 3rd Group
table(genotypes$Group) 
  # maize count: ZMMIL 290, ZMMLR 1256, and ZMMMR 27 = 1573
  # teosinte count: ZMPBA 900, ZMPIL 41, and ZMPJA 34 = 975


dim(snp_position) # 983 rows  15 columns
colnames(snp_position)[1:15] # variable names
# head(snp_position) # 1st SNP_ID, 3rd Chromosome, 4th Position, etc.
table(snp_position$Chromosome) # 6 multiple and 27 unknown
```

### _Data Processing_
`fang_et_al_genotypes.txt` 
First, I am cutting out the groups I need for maize and teosinte.
```{r}
# Filtering `fang_et_al_genotypes.txt` 
maize_genotypes <- genotypes %>% filter(Group == "ZMMIL" |
                                        Group == "ZMMLR" |
                                        Group == "ZMMMR" )
teosinte_genotypes <- genotypes %>% filter(Group == "ZMPBA" |
                                           Group == "ZMPIL" |
                                           Group == "ZMPJA" )
# Checking: both result is 3, which is accurate
n_distinct(maize_genotypes$Group)
n_distinct(teosinte_genotypes$Group) 

```

Then I will transpose the genotype files so that columns become rows by using t() before joining.
```{r}
# MAIZE
t_maize_genotypes <- as.data.frame(t(maize_genotypes))
t_maize_genotypes <- t_maize_genotypes[-c(1,2,3),]
dim(t_maize_genotypes) # 983 rows just like snp_position
# Add Sample_ID back
colnames(t_maize_genotypes) <- maize_genotypes$Sample_ID

# TEOSINTE
t_teosinte_genotypes <- as.data.frame(t(teosinte_genotypes))
t_teosinte_genotypes <- t_teosinte_genotypes[-c(1,2,3),]
dim(t_teosinte_genotypes) # 983 rows just like snp_position
# Add Sample_ID back
colnames(t_teosinte_genotypes) <- teosinte_genotypes$Sample_ID


```

Now I am cleaning snp_position and fixing the join key for genotype files.
Assignment says files formated with 1st column is "SNP_ID", the 2nd column is "Chromosome", the 3rd column is "Position", and subsequent columns are genotype data from either maize or teosinte individuals."
help: `?rownames_to_column` and _https://stackoverflow.com/questions/29511215/how-can-i-convert-row-names-into-the-first-column_ 
```{r}
# Clean snp_position. We only want column 1, 3, 4
cleaned_snp_position <- select(snp_position, 
                               "SNP_ID", "Chromosome", "Position")
# I don't have SNP_ID column name from genotype data so here I add it.
t_maize_genotypes <- tibble::rownames_to_column(t_maize_genotypes, "SNP_ID")
t_teosinte_genotypes <- tibble::rownames_to_column(t_teosinte_genotypes, "SNP_ID")

```

Joining snp_position to both of these genotype files.
help: `?left_join` and _https://r4ds.hadley.nz/joins.html_
```{r}
# MAIZE
joined_maize <- left_join(
  x = t_maize_genotypes, 
  y = cleaned_snp_position, 
  by = "SNP_ID")
tail(names(joined_maize), 5) # it stuck it onto the end, lets move it to the front
joined_maize <- joined_maize %>% select(SNP_ID, Chromosome, Position, everything())


# TEOSINTE
joined_teosinte <- left_join(
  x = t_teosinte_genotypes, 
  y = cleaned_snp_position, 
  by = "SNP_ID")
tail(names(joined_teosinte), 5) # it stuck it onto the end, lets move it to the front
joined_teosinte <- joined_teosinte %>% select(SNP_ID, Chromosome, Position, everything())

```

Now let's check the new datasets : `joined_maize` and `joined_teosinte`
```{r}
unique(joined_maize$Chromosome) # we need to remove multiple and unknown
joined_maize <- filter(joined_maize, 
                Chromosome != "multiple" & Chromosome != "unknown")
unique(joined_maize$Chromosome) # gone!

unique(joined_teosinte$Chromosome) # we need to remove multiple and unknown
joined_teosinte <- filter(joined_teosinte, 
                Chromosome != "multiple" & Chromosome != "unknown")
unique(joined_teosinte$Chromosome)  # gone!
```

Fixing position and removing positions that are not numeric, and hence can not be sorted by increasing position values.
```{r}
class(joined_maize$Position) # character, should be numeric so I don't sort wrong
cleaned_maize <- joined_maize %>%
  mutate(Position = as.numeric(Position)) %>%
  filter(!is.na(Position))

cleaned_teosinte <- joined_teosinte %>%
  mutate(Position = as.numeric(Position)) %>%
  filter(!is.na(Position))

```
This is how I would generate the final files for 1 chromosome.
```{r}

dim(cleaned_maize) # 1576 columns
test_chr_1 <- cleaned_maize %>%
  filter(Chromosome == "1")
test_chr_1_increase <- test_chr_1 %>%
  arrange(Position) %>%
  mutate(across(4:1576, function(x){
    ifelse(x == "?/?", "?", x)
  })
  )

test_chr_1_decrease <- test_chr_1 %>%
  arrange(desc(Position))%>%
  mutate(across(4:1576, function(x){
    ifelse(x == "?/?", "-", x)
  })
  )
``` 

Creating a function.
```{r}

final_files <- function (df, chr, name){
  
  chr_df <- df %>% filter(Chromosome == chr)
  chr_increase <- chr_df %>% 
    arrange(Position) %>%
    mutate(across(-(1:3), function(x){
    ifelse(x == "?/?", "?", x)
  })
  )
  
  chr_decrease <- chr_df %>%
  arrange(desc(Position))%>%
  mutate(across(-(1:3), function(x){
    ifelse(x == "?/?", "-", x)
  })
  )
# write_csv(dataframe, output path should be final_results, delimiters) 
# I need no spaces with it in the final_maize or final_teosinte folder which is inside the final_results folder. 
# set up should be : final_results/final_maize/maize_chr1_increasing.csv
  # which means that the actual file names requires 
    # name + _chr + "chr" (the actual value) + _increasg.csv
   write_csv(chr_increase, paste0("final_results/final_", name, "/", 
                                  name, "_chr", chr, 
                                  "_increasing.csv") )
   write_csv(chr_decrease, paste0("final_results/final_", name, "/", 
                                  name, "_chr", chr, 
                                  "_decreasing.csv") )

}

```

Using `for loop` vs `lapply`.
```{r}
class(cleaned_maize$Chromosome) # characters

for(chr in as.character(1:10)) {
  final_files(cleaned_maize, chr, "maize") 
}

for(chr in as.character(1:10)) {
  final_files(cleaned_teosinte, chr, "teosinte") 
}
```

## Part II Visualization

### _SNPs per chromosome_
What is the distribution of SNPs on and across chromosomes?
```{r}
# Unsure if we want Chromosome with multiple or unknown since they aren't numerically sortable.
# This gives us number of SNP sites on each chromosome

dist_SNPs <- snp_position %>% 
  filter(Chromosome != "multiple" & Chromosome != "unknown") %>%
  mutate(Chromosome = as.numeric(Chromosome)) %>%
  mutate(Chromosome = factor(Chromosome, levels = 1:10)) %>%
  count(Chromosome)
 

  
ggplot(dist_SNPs, aes(x=Chromosome, y=n)) +
  geom_bar(stat = "identity", fill = "lightblue2") + 
  labs (
    title = "Distribution of SNPs on and across chromosomes",
    x = "Chromosome",
    y = "Number of SNPs"
  ) + theme_classic()
```

Are there more SNP positions in maize or teosinte individuals? 
```{r}
maize <-

```

### _Missing data and amount of heterozygosity_
What is the proportion of homozygous and heterozygous sites as well as missing data in each sample and each group?
Hints: Create a new column to indicate whether a particular site is homozygous (has the same nucleotide on both chromosomes (i.e., A/A, C/C, G/G, T/T) or heterozygous (otherwise)). Normalize the height of individual bars using one of the ggplot "position adjustments" options.
```{r}

```


### _Your own visualization_
Visualize one other feature of the dataset. The choice is up to you!
```{r}


```

