---
title: "R_Markdown_Valerie"
author: "Valerie Yu"
date: "`r Sys.Date()`"
output: pdf_document
---

# Valerie's R Assignment
Setting up the R Markdown and loading in R library.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())
library(tidyverse)
library(ggplot2)
library(tidyr)
library(tibble)
```


## Part I
### _Data Inspection_
`fang_et_al_genotypes.txt`

* There are 2782 rows and 986 columns.
* The first 3 columns are `Sample_ID`, `JG_OTU`, and `Group`. Then the rest are SNP data (983).

`snp_position.txt`

* There are 983 rows and 15 columns.
* The columns are named `SNP_ID`, `Chromosome`, `Position`, etc.

```{r}
# Loading in data from "data" folder and looking at column names. 
genotypes <- read_tsv("data/fang_et_al_genotypes.txt")

snp_position <- read_tsv("data/snp_position.txt")



# Inspecting the data
dim(genotypes) # 2782 rows  986 columns
colnames(genotypes)[1:15] # variable names 
# head(genotypes) # 1st Sample_ID, 3rd Group
table(genotypes$Group) 
  # maize count: ZMMIL 290, ZMMLR 1256, and ZMMMR 27 = 1573
  # teosinte count: ZMPBA 900, ZMPIL 41, and ZMPJA 34 = 975


dim(snp_position) # 983 rows  15 columns
colnames(snp_position)[1:15] # variable names
head(snp_position) # 1st SNP_ID, 3rd Chromosome, 4th Position, etc.
table(snp_position$Chromosome) # 6 multiple and 27 unknown
```

### _Data Processing_
`fang_et_al_genotypes.txt` 
First, I am cutting out the groups I need for maize and teosinte.
```{r}
# Filtering `fang_et_al_genotypes.txt` 
maize_genotypes <- genotypes %>% filter(Group == "ZMMIL" |
                                        Group == "ZMMLR" |
                                        Group == "ZMMMR" )
teosinte_genotypes <- genotypes %>% filter(Group == "ZMPBA" |
                                           Group == "ZMPIL" |
                                           Group == "ZMPJA" )
# Checking: both result is 3, which is accurate
n_distinct(maize_genotypes$Group)
n_distinct(teosinte_genotypes$Group) 
```

Then I will transpose the genotype files so that columns become rows by using t() before joining.
```{r}
# MAIZE
t_maize_genotypes <- as.data.frame(t(maize_genotypes))
t_maize_genotypes <- t_maize_genotypes[-c(1,2,3),]
dim(t_maize_genotypes) # 983 rows just like snp_position

# TEOSINTE
t_teosinte_genotypes <- as.data.frame(t(teosinte_genotypes))
t_teosinte_genotypes <- t_teosinte_genotypes[-c(1,2,3),]
dim(t_teosinte_genotypes) # 983 rows just like snp_position
```

Now I am cleaning snp_position and fixing the join key for genotype files.
Assignment says files formated with 1st column is "SNP_ID", the 2nd column is "Chromosome", the 3rd column is "Position", and subsequent columns are genotype data from either maize or teosinte individuals."
help: `?rownames_to_column` and _https://stackoverflow.com/questions/29511215/how-can-i-convert-row-names-into-the-first-column_ 
```{r}
# Clean snp_position. We only want column 1, 3, 4
cleaned_snp_position <- select(snp_position, 
                               "SNP_ID", "Chromosome", "Position")
# I don't have SNP_ID column name from genotype data so here I add it.
t_maize_genotypes <- tibble::rownames_to_column(t_maize_genotypes, "SNP_ID")
t_teosinte_genotypes <- tibble::rownames_to_column(t_teosinte_genotypes, "SNP_ID")
```

Joining snp_position to both of these genotype files.
help: `?left_join` and _https://r4ds.hadley.nz/joins.html_
```{r}
# MAIZE
joined_maize <- left_join(
  x = t_maize_genotypes, 
  y = cleaned_snp_position, 
  by = "SNP_ID")
tail(names(joined_maize), 5) # it stuck it onto the end, lets move it to the front
joined_maize <- joined_maize %>% select(SNP_ID, Chromosome, Position, everything())


# TEOSINTE
joined_teosinte <- left_join(
  x = t_teosinte_genotypes, 
  y = cleaned_snp_position, 
  by = "SNP_ID")
tail(names(joined_teosinte), 5) # it stuck it onto the end, lets move it to the front
joined_teosinte <- joined_teosinte %>% select(SNP_ID, Chromosome, Position, everything())


```



For maize (Group = ZMMIL, ZMMLR, and ZMMMR in the third column of the fang_et_al_genotypes.txt file) we want 20 files in total:
```{r}

```

For teosinte (Group = ZMPBA, ZMPIL, and ZMPJA in the third column of the fang_et_al_genotypes.txt file) we want 20 files in total:
```{r}

```

## Part II Visualization

### _SNPs per chromosome_
What is the distribution of SNPs on and across chromosomes?
Are there more SNP positions in maize or teosinte individuals?

### _Missing data and amount of heterozygosity_
What is the proportion of homozygous and heterozygous sites as well as missing data in each sample and each group?  

### _Your own visualization_
Visualize one other feature of the dataset. The choice is up to you!